---
title: "RHRVEasy tutorial"
output: 
  pdf_document: 
    toc: yes
  html_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

RHRVEasy automates all steps of a Heart Rate Variability (HRV) analysis, 
including data processing, indices calculation, and statistical analysis. It 
takes as input a list of folders, each containing the recordings of a same 
population. It calculates time, frequency, and nonlinear domain HRV indices, 
and then it applies hypothesis test, and corrects the significance levels. If 
there are more than two experimental groups and statistically significant 
differences are found, it performs a post-hoc analysis to find out which groups 
have the differences. 

# 0. Set up required to run this tutorial
This tutorial uses the recordings of the [Normal Sinus Rhythm RR Interval 
Database](https://physionet.org/content/nsr2db/1.0.0/) (hereinafter referred to 
as NSR_DB) and of the [Congestive Heart Failure RR Interval 
Database](https://archive.physionet.org/physiobank/database/chf2db/) 
(hereinafter referred to as CHF_DB). The first is made of healthy subjects, and 
the second of patients with a severe cardiac pathology. Hence significant 
differences in many HRV indices are expected between them.

To illustrate how to work with more than two experimental groups, half of the 
recordings from both databases were randomly selected. The NSR_HALF_DB database 
is made up of 24 recordings randomly chosen from the 58 of the NSR database. 
The CHF_HALF_DB database is made up of 14 recordings chosen from the 28 of the 
CHF database. The four databases are available in the [RHRVEasy GitHub 
repository](https://github.com/constantino-garcia/RHRVEasy). To run this tutorial, 
the following four variables must contain the folder of the local machine where 
the files of each of these databases are. Note that the four folders can be found 
as a zip file in the Github repository, under the `RRData` directory (refer to the `data` 
section in the GitHub `README`).

```{r}
library("RHRVEasy")

basePath <- "RRData/" 
NSR_DB <- file.path(basePath, "normal")
CHF_DB <- file.path(basePath, "chf")
NSR_HALF_DB <- file.path(basePath, "normal_half")
CHF_HALF_DB <- file.path(basePath, "chf_half")
```

RHRVEasy permits creating an Excel spreadsheet with all the HRV indices 
calculated for each recording. The following variable must contain the folder 
on the local machine where the Excel spreadsheet is to be saved:

```{r}
spreadsheetPath <- basePath
```

# 1. Time and frequency analysis

`RHRVEasy` enables the user to carry out a full HRV analysis by just invoking a 
function with a single mandatory parameter: a list with the folders containing 
the recordings of the experimental groups. This list must have at least two 
folders. Each folder must contain all the RR recordings of the same 
experimental group and no additional files, as `RHRVEasy` will try to open all 
the files in those folders. The name that will be used to refer to each 
experimental group within `RHRVEasy` will be the name of the folder in which its 
recordings are located.

The following function call computes the time and frequency indices for the 
NSR_DB and CHF_DB databases, and and performs a statistical comparison of each 
index correcting the significance level with the Bonferroni method. Note the 
use of the `nJobs` to use several cores and parallelize the computations. Whith `nJobs = -1`, it uses all available cores; if an integer greater than 0 is indicated, it uses the number of cores indicated by the integer.

```{r, results=FALSE, cache=2}
easyAnalysis <- RHRVEasy(folders = c(NSR_DB, CHF_DB), nJobs = -1)
```

When the returned object is displayed in the console, it shows which indices 
present statistically significant differences:

```{r}
print(easyAnalysis)
```
All computed indices, as well as all p-values resulting from all comparisons, 
are stored in `data.frames` contained in the object. Two different sets of 
p-values are available; the ones obtained before (`p.value`) and after 
(`adj.p.value`) applying the significance level correction:

```{r}
# HRVIndices
head(easyAnalysis$HRVIndices)

# Statistical analysis
head(easyAnalysis$stats)
```

The `format` parameter specifies the format in which the RR intervals are 
stored. All formats supported by the RHRV package can be used: `WFDB`, `ASCII`, 
`RR`, `Polar`, `Suunto`, `EDFPlus` or `Ambit` (check the [RHRV 
website](https://rhrv.r-forge.r-project.org/) for more information). The 
default format is RR, where the beat distances in seconds are stored in a 
single column of an ASCII file. This is the format of the four databases used 
in this tutorial.

By default, the frequency analysis is performed using the Fourier transform. It 
is also possible to use the Wavelet transform pasing the value `'wavelet'` to 
the `typeAnalysis` parameter (check the paper "García, C. A., Otero, A., Vila, 
X., & Márquez, D. G. (2013). A new algorithm for wavelet-based heart rate 
variability analysis. Biomedical Signal Processing and Control, 8(6), 542-550" 
for details):

```{r, results=FALSE, cache=2}
easyAnalysisWavelet <- RHRVEasy(
  folders = c(NSR_DB, CHF_DB), 
  typeAnalysis = 'wavelet', 
  n_jobs = -1
)
```

The results are similar to the previous ones:

```{r}
print(easyAnalysisWavelet)
```


# 2. Correction of the significance level

Given that multiple statistical tests are performed on several HRV indices, a 
correction of the significance level should be applied. The Bonferroni method 
is used by default. This behavior can be overridden with 
the parameter `correctionMethod` of `RHRVEasy`. The possible values of this 
parameter besides  `bonferroni` are `holm`, `hochberg`, `hommel`,
`BH` (Benjamini & Hochberg),  `fdr` (false discovery rate), 
`BY` (Benjamini & Yekutieli), and `none`  (indicating that no correction is to
be made). Furthermore, there is no need to recompute the HRV indices to apply 
a different correction method, but the `RHRVEasyStats` function can be used to this end. The confidence
level can also be changed using the `significance` parameter (in both 
`RHRVEasy` and `RHRVEasyStats` functions).

```{r}
easyAnalysisFDR <- RHRVEasyStats(easyAnalysis, correctionMethod =  'fdr')
pValues <- merge(
  easyAnalysis$stats, 
  easyAnalysisFDR$stats,
  by = setdiff(names(easyAnalysis$stats), "adj.p.value"),
  suffixes = c(".bonf", ".fdr")
)
#Let us compare the p-values obtained with different correction methods 
print(
  head(
    pValues[, c("HRVIndex", "p.value", "adj.p.value.bonf", "adj.p.value.fdr")]
  )
) 
```


# 3. Saving the indices to an Excel spreadsheet

If the argument `saveHRVindicesInPath` is specified when invoking the function 
`RHRVEasy`, an Excel spreadsheet with all the HRV indices calculated for each 
recording will be created in the path specified specified in this parameter. 
The name of the spreadsheet generated is "<group 1 name>_Vs_<group 2 name> 
.xlsx":

```{r, eval=FALSE}
easyAnalysis <- RHRVEasy(folders = c(NSR_DB, CHF_DB), 
                         saveHRVIndicesInPath = spreadsheetPath)
```

This spreadsheet can also be generated from the object returned by `RHRVEasy` 
by calling the function `saveHRVIndices`.

```{r}
saveHRVIndices(easyAnalysis,saveHRVIndicesInPath = spreadsheetPath)
```

# 4. Comparing more than two experimental groups

If the analysis involves three or more groups, when statistically significant 
differences are found among them it does not necessarily mean that there are 
statistically significant differences between all pairs of groups. In such a 
scenario the Dunn post-hoc is used to find which pairs of groups present 
differences:

```{r, cache=2, results=FALSE}
#Comparison of the four databases
easyAnalysis4 <- RHRVEasy(
  folders = c(NSR_DB, CHF_DB, NSR_HALF_DB, CHF_HALF_DB),
  nJobs = -1
)
```
```{r}
print(easyAnalysis4)
```

Note that the `stats` `data.frame` now contains a column named `pairwise` storing
the  results of the post-hoc analysis for those indices where the omnibus test 
has been significant:

```{r}
print(head(easyAnalysis4$stats))
# Let's print the post-hoc comparisons for "SDNN"
print(head(easyAnalysis4$stats$pairwise[[1]]))
```


# 5. Overwriting default parameters

Any parameter of any RHRV function can be specified as an additional parameter 
of the `RHRVEasy` function; in this case, the default value used for that 
parameter will be overwritten by the one specified for the user. The default 
values used in the `RHRVEasy` package are the same as those used in the RHRV 
package. For more information about the parameters available you can consult 
the [RHRV website](https://rhrv.r-forge.r-project.org/). For example, the 
following analysis modifies the the limits of the ULF, VLF, LF and HF spectral 
bands, and uses an interpolation frequency (`freqhr`) of 2 Hz:

```{r, results=FALSE, eval=FALSE}
easyAnalysisOverwritten <- RHRVEasy(folders = c(NSR_DB, CHF_DB),
                                    freqhr = 2, 
                                    ULFmin = 0, ULFmax = 0.02, 
                                    VLFmin = 0.02,  VLFmax = 0.07, 
                                    LFmin = 0.07, LFmax = 0.20, 
                                    HFmin = 0.20, HFmax = 0.5)
```

# 6. Nonlinear analysis

The calculation of the nonlinear indices requires considerable computational 
resources, specially the Recurrence Quantification Analysis (RQA). 
Whereas in a typical HRV analysis the computation of all the time 
and frequency domain indices for a few dozens of recordings often completes 
within a few minutes, the computation of the nonlinear indices could last many 
hours. That's why the boolean parameters `nonLinear` and `doRQA` are set to `FALSE`
by default. If these parameters are not change, only time and frequency indices
will be calculated, as in the previous sections.

**Warning**: the following sentence, will take  several hours to execute on a 
medium to high performance PC. You may reproduce the results of the paper by 
running this chunk of code. Alternatively, you may load the RDS file 
`RRData/paperExperiments.RDS` to explore the statistical results.

```{r, eval=FALSE}
fullAnalysis <- RHRVEasy(
  folders = c(NSR_DB, CHF_DB, NSR_HALF_DB, CHF_HALF_DB),
  nJobs = -1,
  nonLinear =  TRUE, 
  doRQA = TRUE,
  saveHRVIndicesInPath = spreadsheetPath
)  
```
